Towards open collaboration in biomedical and health research: An open collaborative scoping review

### 2025-02-10 doc/paper.qmd
updated results section and started to add reference to the bib-file

### Scoping review identifies only 11 papers discussing open collaboration

### Practices of open collaboration

Across the papers found in our literature search, X out of 11 discuss
actual experiences on implementing open collaboration, albeit in
different organizational levels. These range from lab level (Turoman et
al) center level (Bush et al and Grange et al), multi-center level
(ManyPrimates et al and Grange et al) and field level: psychology
(Alessandorni et al), Spinal Cord injury (Torres-Espin et al) and
neuroimaging (Niso et al). All of these share in common the need for an
infrastructure that promotes open science and open collaboration via
starting building a community of collaborators or making tools (usually
online resources) that researchers can use to reach FAIR (Finndable,
Accessible, Interoperable and Reusable) goals while doing research.
Before delving in those, here we provide the specific needs for
implementing open collaboration in each organizational level, as well as
describing the differences in approaches among experiences between and
across levels (Figure 2).

#### Implementation of open collaboration practices at laboratory level

#### Implementation of open collaboration practices at center level

For center-level only one paper was identified (Bush et al), providing
the example of a center of neuroimaging describing first how they
identified the needs for open collaboration, then the individual-level
practices that researchers should adapt and then the center-level
changes to promote reaching FAIR goals.

#### Implementation of open collaboration practices at multi-center level

Interestingly, the level of organization with most papers is the
multi-center level with two papers from ManyPrimates et al (ManyPrimates
2019 and ManyPrimates 2021) and Grange et al. In this organization level
the most urgent needs are: 1) identifying the needs for open
collaboration and 2) creating an infrastructure that allows those needs
to be satisfied. Indeed, Grange et al describes a systematic review as
an exploration of the needs for 11 UKRN centers to implement open
collaborative practives. In this paper they specially focus on tools and
resources among the 11 institutions and make clear the need for starting
to utilize open resources as a first step to adapt open collaborative
practices. The first paper from ManyPrimates et al \[REF\] describes the
start of another multi-center collaboration, albeit from different
institutions. It is interesting to observe that in this situation, the
complex issue is to start the collaboration itself, which was ignited
thanks to a symposium and an email chain among future collaborators. In
contrast, ManyPrimates could quickly develop an open collaborative
structure with relative ease in comparison to changing an already solid
infrastructure, like in the case of Grange et al.

#### Implementation of open collaboration practices at field level

@@

## Discussions

## Conclusions

>>>>

```{r}
library(targets)

n_papers <- tar_read(records_title_only) %>% nrow() %>% as.numeric()

n_papers_title_review <- tar_read(records_after_title_exclusion) %>% nrow() %>% as.numeric()

n_papers_excluded_title_review <- n_papers - n_papers_title_review

n_titles_selected <- tar_read(titles_selected) %>% nrow() %>% as.numeric()

n_paper_full_text <- tar_read(abstracts_selected) %>% nrow() %>% as.numeric()

n_papers_selected_for_review <- read.csv(here("data", "article_characteristics.csv"), header = TRUE) %>% nrow() %>% as.numeric()

#get table of study characteristics 
article_characteristics <- read.csv(here("data", "article_characteristics.csv"))

n_guides <- article_characteristics %>% filter(paper_type == "guide") %>% nrow() %>% as.numeric()

n_examples <- article_characteristics %>% filter(paper_type == "example") %>% nrow() %>% as.numeric()

n_surveys <- article_characteristics %>% filter(paper_type == "survey") %>% nrow() %>% as.numeric()

n_org <- article_characteristics %>% filter(str_detect(themes, "organization")) %>% nrow() %>% as.numeric()

n_tools <- article_characteristics %>% filter(str_detect(themes, "tools")) %>% nrow() %>% as.numeric()

n_education <- article_characteristics %>% filter(str_detect(themes, "education")) %>% nrow() %>% as.numeric()

n_researcher <- article_characteristics %>% filter(relevant_for == "researcher") %>% nrow() %>% as.numeric()

n_research_group <- article_characteristics %>% filter(relevant_for == "research group") %>% nrow() %>% as.numeric()

n_research_center <- article_characteristics %>% filter(relevant_for == "research center") %>% nrow() %>% as.numeric()

n_research_centers <- article_characteristics %>% filter(relevant_for == "multiple centers") %>% nrow() %>% as.numeric()

n_research_field <- article_characteristics %>% filter(relevant_for == "research field") %>% nrow() %>% as.numeric()

```

The search yielded a total of `r n_papers` papers from which we filtered
`r n_papers_excluded_title_review` based on key words. The remaining
`r n_papers_title_review` titles were independently screened by three
reviewers from which `r n_titles_selected` abstracts were screened. Of
these, `r n_paper_full_text` papers were assessed in full-text and
`r n_papers_selected_for_review` were included in the scoping review.

The identified papers included `r n_guides` guides, `r n_examples`
examples and `r n_surveys` surveys.

`r n_org` of the `r n_papers_selected_for_review` papers included
recommendations for organizational practices to improve open
collaboration, `r n_tools` mentioned specific tools used for open
collaboration and `r n_education` highlighted educational needs. The
identified papers discussed open collaboration on different
organizational levels. `r n_researcher` were mostly directed at the
individual researcher, `r n_research_group` at a research group,
`r n_research_center` at a research center, `r n_research_centers` at
several research centers and `r n_research_field` at entire research
fields.

### Organizational practices of open collaboration

A common theme across organizational levels was the need for
infrastructure that promotes open collaboration, building a community of
collaborators and using online tools in accordance with the FAIR
(Finndable, Accessible, Interoperable and Reusable) principles.

individual level - add text

laboratory/research group level - add text

For center-level one paper described how they transitioned their
neuroimaging research center toward more open and collaborative
practices. Their central focus was on creating more open code and data
to improve open collaboration @bush_lessons_2022. First, they used
open-source code, then standardized data formats, developed analysis
pipelines, shared data openly and developed protocols for data
documentation. While the organizational transition to improve open
collaboration was described, little was mentioned about how this has
improved collaborative work at the research center.

Interestingly, the level of organization with most papers was the
multi-center level with two papers from ManyPrimates et al
(@manyprimates_collaborative_2019, @manyprimates_collaboration_2021) and
Grange et al. (@grange_ukrn_2023). In this organization level the most
urgent needs are: 1) identifying the needs for open collaboration and 2)
creating an infrastructure that allows those needs to be satisfied.
Indeed, Grange et al describes a systematic review as an exploration of
the needs for 11 UKRN centers to implement open collaborative practives.
In this paper they specially focus on tools and resources among the 11
institutions and make clear the need for starting to utilize open
resources as a first step to adapt open collaborative practices. The
first paper from ManyPrimates et al @manyprimates_collaborative_2019
describes the start of another multi-center collaboration, albeit from
different institutions. It is interesting to observe that in this
situation, the complex issue is to start the collaboration itself, which
was ignited thanks to a symposium and an email chain among future
collaborators. In contrast, ManyPrimates could quickly develop an open
collaborative structure with relative ease in comparison to changing an
already solid infrastructure, like in the case of Grange et al.

research field level

Here, we provide the specific needs for implementing open collaboration
in each organizational level, as well as describing the differences in
approaches among experiences between and across levels (Figure 2).

@@

### Educational needs

## Discussion

### Conclusions
